Spirals launched by a planet in a shadowed disk

### 2024-04-15 src/tex/ms.tex
finish edits

\section{Methods}
\label{sec:methods}
\subsection{Simulation setup}
\label{subsec:setup}
We use the magneto-hydrodynamics code Athena++\footnote{\url{https://github.com/PrincetonUniversity/athena}} 
\citep{stone2020} version 21.0 \citep{athena++developmentteam2021} to simulate a 2D gas disk. The simulation was set up in cylindrical coordinates with 149
cells in the $\ln{r}$ direction with $r \in [0.4, 2.5]$ and 512 cells in the $\phi \in [-\pi, \pi]$ direction. The inner and outer boundaries have a user-defined fixed boundary
condition and the azimuthal boundary has a periodic boundary condition to ensure material can orbit the central star indefinitly. The dimensionality of the problem
is set up so that all masses are in $M_\odot$, distances are in AU, and times are in year/$2\pi$. This means that a planet placed in a circular orbit at $r=1$
will have an orbital period of $2\pi$. We chose a planet mass of $\mpl$ in order to create a detectable effect in the presence of noise from the shadow and
to not stray too far from the linear spiral regime.

The disk is set up so that $\rho = 1$ at $r=1$ and has the radial profile $\rho \propto r^{-1}$. The gas in the disk initially moves at the local keplerian velocity.
A shadow on the disk is prescribed by setting the sound speed -- analogous to setting the temperature as 
\begin{equation}
    c_s = \frac{\partial P}{\partial \rho} \propto T
\end{equation}
We ensure that the shadow does not move with the gas in its orbit by setting the cooling timescale to $10^{-6} \ll 1$. Essentially, the temperature is reset each time
the grid is updated.

@@

\begin{equation*}
    \label{eq:H}
    \begin{array}{cccc}
        H = & h & \text{if} & |\sin{\phi}| \le \sin{a}  \\
            & 1 & \text{if} & |\cos{\phi}| \le \sin{b} \\
            & h + \frac{1-h}{2} \left (  1-\cos{\left ( \frac{2 \pi | \phi + m\pi |}{\pi - 2a - 2b}  \right )  } \right ) & & \text{otherwise}
    \end{array}
\end{equation*}
and
\begin{equation}
    \begin{array}{cccc}
        m = & 1 & \text{if} & \phi < -\frac{\pi}{2} \\
            & -1 & \text{if} & \phi > \frac{\pi}{2} \\
            & 0 & & \text{otherwise}
    \end{array}
\end{equation}
Figure \ref{fig:cs} shows the sound speed profiles used in this work.

@@

\begin{equation}
    \zeta = \arctan |(\Omega_{\rm pat} - \Omega)r/c_{\rm s}|
\end{equation}

@@

where we have removed the log from the differential and divided out the additional factor of $r$. We can separate the differential to see
\begin{equation}
    \int_{\phi_p}^{\phi_f} c_{\rm s} d \phi = \int_{1}^{r_f} (1-r^{-3/2}) d r
    \label{eq:analytic_shadow}
\end{equation}

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/spiral_shape_analytic.pdf}
    \caption{
        Analytic spiral curves for a variety of situations. Here we consider both the narrow and wide cases for multiple values of $h$.
        The $h=0.9$ case which we simulate numerically very nearly matches the shadowless case, shown in black. In order to get
        variations similar to those seen in our simulations we have to set $h$ to be much less than 1. The region near the planet
        has been cut out because the analytic treatment is not valid there.
    }
    \label{fig:analytic_spiral}
    \script{plot_sound_speed_analytic.py}
\end{figure}

@@

It can be clearly seen in Figures \ref{fig:end}, \ref{fig:sp_narrow} \& \ref{fig:sp_wide} that the morphology of the disk is dominated
by effects from the shadow, even for the very weak shadows we simulate in this work. In all simulations involving a shadow we observe a very clear
ring pattern. These rings appear after $t\sim10~P_{\rm orb}$ and are static, as shown in Figure \ref{fig:slice} It is interesting to note that
the innermost ring in the narrow case is a minimum, while in the wide case it is a maximum. Inside the innermost ring is a spiral region where gas
is interacting with the boundary.

@@

We also leave much of the shadow parameter space unexplored. Future investigations could study a variety of parameters, including
\begin{enumerate}
    \item The properties of the gas. We assume instant heating and cooling, but a longer timescale would mean that $c_{\rm s}$ is not
        purely a function of $\phi$ as gas in the inner portions of the disk would travel a significant fraction of its orbit before cooling completely.
    \item The radial temperature profile. We assume that $c_{\rm s}$ is not a function of $r$. However, the outer portions of the disk would realistically
        be cooler.
    \item Shadow shape. We assume shadows cast by a polar-inclined interior disk. However, if the misalignment is small then shadows on a disk can have a very different
        shape -- one that is a function of both $\phi$ and $r$.
    \item Motion of the shadow. In this study we simulated a static shadow -- analogous to that produced by an inclined interior disk. However,
        if instead a shadow is caused by an eclipsing interior binary then it would sweep around the disk on the timescale of the binary orbit.
    \item Limiting regimes of $h$. At what value do we start to see rings formed? What happens in the case of a very strong shadow?
    \item Different width regimes. At what point does the interior ring switch from a minimum to a maximum?
\end{enumerate}

>>>>

\section{Methods}
\label{sec:methods}
\subsection{Simulation setup}
\label{subsec:setup}
We use the magneto-hydrodynamics code Athena++\footnote{\url{https://github.com/PrincetonUniversity/athena}} 
\citep{stone2020} version 21.0 \citep{athena++developmentteam2021} to simulate a 2D gas disk. The simulation was set up in cylindrical coordinates with 149
cells in the $\ln{r}$ direction with $r \in [0.4, 2.5]$ and 512 cells in the $\phi \in [-\pi, \pi]$ direction. The inner and outer boundaries have a user-defined fixed boundary
condition and the azimuthal boundary has a periodic boundary condition to ensure material can orbit the central star indefinitly. The dimensionality of the problem
is set up so that all masses are in $M_\odot$, distances are in AU, and times are in year/$2\pi$. This means that a planet placed in a circular orbit at $r=1$
will have an orbital period of $2\pi$. We chose a planet mass of $\mpl$ in order to stay in the linear spiral regime.

The disk is set up so that $\rho = 1$ at $r=1$ and has the radial profile $\rho \propto r^{-1}$. The gas in the disk initially moves at the local keplerian velocity.
A shadow on the disk is prescribed by setting the sound speed -- analogous to setting the temperature as 
\begin{equation}
    c_s = \sqrt{\frac{\partial P}{\partial \rho}} \propto \sqrt{T}
\end{equation}
We ensure that the shadow does not move with the gas in its orbit by setting the cooling timescale to $10^{-6} \ll 1$. Essentially, the temperature is reset each time
the grid is updated.

@@

\begin{equation*}    
    \begin{array}{cccc}
        H = & h & \text{if} & |\sin{\phi}| \le \sin{a}  \\
            & 1 & \text{if} & |\cos{\phi}| \le \sin{b} \\
            & h + \frac{1-h}{2} \left (  1-\cos{\left ( \frac{2 \pi | \phi + m\pi |}{\pi - 2a - 2b}  \right )  } \right ) & & \text{otherwise}
    \end{array}
\end{equation*}
and
\begin{equation}
    \begin{array}{cccc}
        m = & 1 & \text{if} & \phi < -\frac{\pi}{2} \\
            & -1 & \text{if} & \phi > \frac{\pi}{2} \\
            & 0 & & \text{otherwise}
    \end{array}
    \label{eq:H}
\end{equation}
Figure \ref{fig:cs} shows the sound speed profiles used in this work.

@@

\begin{equation}
    \zeta = \text{arccot}~|(\Omega_{\rm pat} - \Omega)r/c_{\rm s}|
\end{equation}

@@

where we have removed the log from the differential and divided out the additional factor of $r$. We can separate the differential to see
\begin{equation}
    \int_{\phi_p}^{\phi_f} c_{\rm s} d \phi = \int_{1}^{r_f} |1-r^{-3/2}| d r
    \label{eq:analytic_shadow}
\end{equation}

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/spiral_shape_analytic.pdf}
    \caption{
        Analytic spiral solutions for a variety of shadows. Here we consider both the narrow and wide cases for multiple values of $h$.
        The $h=0.9$ case which we simulate numerically very nearly matches the shadowless case, shown in black. In order to get
        variations similar to those seen in our simulations we have to set $h$ to be much less than 1. The region near the planet
        has been cut out because the analytic treatment is not valid there. {\bf Left:} Location of spiral peaks. {\bf Right:}
        $\zeta$ as a function of radius. Note that the pitch angle here never exceeds the pitch angle in the control case.
    }
    \label{fig:analytic_spiral}
    \script{plot_sound_speed_analytic.py}
\end{figure*}

@@

It can be clearly seen in Figures \ref{fig:end}, \ref{fig:sp_narrow} \& \ref{fig:sp_wide} that the morphology of the disk is dominated
by effects from the shadow, even for the very weak shadows we simulate in this work. In all simulations involving a shadow we observe a very clear
ring pattern. These rings appear after $t\sim10~P_{\rm orb}$ and are static, as shown in Figure \ref{fig:slice}. It is interesting to note that
the innermost ring in the narrow case is a maximum, while in the wide case it is a minimum. Inside the innermost ring is a spiral region where gas
is interacting with the boundary.

@@

We also leave much of the shadow parameter space unexplored. Future investigations could study a variety of parameters, including
\begin{enumerate}
    \item The properties of the gas. We assume instant heating and cooling, but a longer timescale would mean that $c_{\rm s}$ is not
        purely a function of $\phi$ as gas in the inner portions of the disk would travel a significant fraction of its orbit before cooling completely.
    \item The radial temperature profile. We assume that $c_{\rm s}$ is not a function of $r$. However, the outer portions of the disk would realistically
        be cooler.
    \item Shadow shape. We assume shadows cast by a polar-inclined interior disk. However, if the misalignment is small then shadows on a disk can have a very different
        shape -- one that is a function of both $\phi$ and $r$.
    \item Motion of the shadow. In this study we simulated a static shadow -- analogous to that produced by an inclined interior disk. However,
        if instead a shadow is caused by an eclipsing interior binary then it would sweep around the disk on the timescale of the binary orbit.
    \item Limiting regimes of $h$. At what value do we start to see rings formed? What happens in the case of a very strong shadow?
    \item Different width regimes. At what point does the interior ring switch from a minimum to a maximum?
    \item Convergence to a steady state. The timescales investigated in this work are very short and a simulation
        that covers another order of magnitude in time could uncover new behavior.
\end{enumerate}
